```{r, echo = FALSE}
suppressPackageStartupMessages(library("foehnix"))
```

## Foehnix Demo

This page shows a _real world_ demo of the automated foehn classification
method **foehnix** based on meteorological observations from two sites in
Tyrol, Austria. [One station](reference/ellboegen.html) is located in the
Brenner valley south of Innsbruck on 1080 meters above mean sea level, the
[second station](reference/sattelberg.html) is located close to the main alpine
ridge on 2107 meters above mean sea level.

```{r, echo = TRUE}
data("ellboegen"); data("sattelberg")
head(ellboegen);   head(sattelberg)
```

This example shows how to perform foehn classification for
[Ellboegen](reference/ellboegen.html) using the information of the
[Sattelberg](reference/sattelberg.html) station as additional source of
information.  The [Sattelberg](reference/sattelberg.html) station is located
south of [Ellboegen](reference/ellboegen.html) close to the alpine crest and
thus, during foehn events, measuring the upstream air mass descending down on
the north side of the Alps.

### South Foehn Classification for Ellboegen

To perform the classification we first of all have to prepare our data sets.
`foehnix(...)` expects the input to be of class `zoo` (zoo time series object;
based on the \emph{R} `zoo` package). Thus, after loading our example data
sets, we first have to convert the `data.frame`s into valid `zoo` objects:

```{r}
# Loading observations (data.frame), convert to zoo
data("ellboegen", package = "foehnix")
data("sattelberg", package = "foehnix")
timefun    <- function(x) as.POSIXct(x,  origin = "1970-01-01", tz = "UTC")
ellboegen  <- zoo(ellboegen[,-1],  timefun(ellboegen$timestamp))
sattelberg <- zoo(sattelberg[,-1], timefun(sattelberg$timestamp))
```

As we want to use information from both stations: rename the variables of the
`sattelberg` object and combine the two data sets into one `zoo` object called
`data`.

```{r}
# Modify sattelberg variable names (crest_ identifies Sattelberg
# observations, our crest station) and combine both data sets.
names(sattelberg) <- paste0("crest_", names(sattelberg))
data <- merge(ellboegen, sattelberg)

# NOTE: for demonstration purpose we split the data set into
# two pieces: a training and a test data set. 
# The training data set is used to estimate the mixture model, the
# test data set will be used for "predictions" (see "Predictions"
# section of this file).
train <- head(data, nrow(data) - 10)
test  <- tail(data, 10)
rm(data) # Delete 'data'
```

In addition we are calculating the potential temperature difference
between the two sites using a dry adiabatic lapse rate of
one 10 degrees Celsius per 1000 meters. The difference in altitude
between the two stations is 1027 meters which yields:

```{r}
# Dry adiabatic temperature difference between 
# Sattelberg (data$crest_t) and Ellboegen (data$t) corrected by
# 1027/10 degrees.
train$diff_t <- train$crest_t + 10.27 - train$t
```

South foehn flow at the valley station in [Ellboegen](reference/ellboegen.html)
has a wind direction of about 133 degrees. We are using a (relatively weak)
wind direction filter with a wind sector of +/-90 degrees around 133 degrees.
For the classification only observations/times are used where the corresponding
wind direction (`dd`) lies within the wind sector `>=43` and `<= 223` (180
degrees sector).  In addition we can make use of the information provided by
the crest station ([Sattelberg](reference/sattelberg.html)) in several ways:

1. We use the temperature difference between the two
   stations as main variable for the classification (`diff_t`).
2. define wind sector filter at the target station ([Ellboegen](reference/ellboegen.html)).
   Only use observations with a wind direction (`dd`) inside the
   wind sector of `>=43` and `<=223` degrees for classification
   (`dd = c(43, 223)`).
2. specify a second filter such that only observations are
   used for classification where the crest station shows
   a southerly flow (wind direction within `>= 90` and `<= 270`,
   `sat_dd = c(90, 270)`).

```{r, echo = TRUE, results = 'hide'}
# Specify our wind filter rule(s)
filter <- list(dd = c(43, 223), crest_dd = c(90, 270))

# Estimate the mixture model for automated foehn classification
mod <- foehnix(diff_t ~ ff, data = train, filter = filter, switch = TRUE)
```

## Check the Model

A set of S3 methods for object of class `foehnix` are available, This includes
methods to access information criteria, methods to check the estimated model
(model settings), methods to return the estimated coefficients and
probabilities, and some default plots for model assessment.

Let's check our demo model `mod` estimated above:

```{r}
# Information criteria
# logLik: final log-likelihood sum of the model
# edf: effective degrees of freedom
# AIC: Akaike information criterion
# BIC: Bayesian information criterion
c(logLik(mod), edf(mod), AIC(mod), BIC(mod))
```

The `print` and `summary` methods show some information about the model such as
number of observations used for the classification, the filter and its effect,
and the corresponding information criteria.

```{r}
print(mod)
```

The `coef` or `coefficients` method returns the estimated
coefficients.

```{r}
coef(mod)
c <- coef(mod)
print(c["mu1"])
print(c["mu2"])
```

The method `plot` provides a set of model assessment plots.  Different types
are available, if no `which` argument is set all will be shown one after
another.

`which = "loglik"` shows the log-likelihood path of the two components of the
model (the part for the two components of the mixture model plus the one of the
concomitant model) plus the full likelihood path.  The x-axis shows the
log-iterations of the EM algorithm (see [statistical
models](articles/statmodel.html) for more details).  Increasing values indicate
positive log-likelihood contributions (improvement of the model).


```{r, fig = TRUE, fig.width = 10, fig.height = 6}
plot(mod, which = "loglik", log = F)
```

`which = "loglikcontribution"` shows the same information as `which = "loglik"`
but with respect to the log-likelihood of the initial parameters. Increasing
values indicate positive log-likelihood contributions (improvement of the
model).

```{r, fig = TRUE, fig.width = 10, fig.height = 6}
plot(mod, which = "loglikcontribution")
```

Similar to the plots above we can also check the paths of the coefficients
trough the iterations of the EM algorithm with `which = "coef"`:

```{r, fig = TRUE, fig.width = 10, fig.height = 6}
plot(mod, which = "coef")
```

Last but not least the `which = "hist"` plot shows the conditional histograms
of the two clusters. Two histograms are shown given the main covariate of the
`foehnix` mixture model, in this case (model `mod`) it is `delta_t`, the
potential temperature difference between the valley station
[Ellboegen](reference/ellboegen.html) and the crest station
[Sattelberg](reference/sattelberg.html).  The left panel corresponds to the "no
foehn" component or cluster, the right one to the "foehn" component. If a
`filter` rule is used: only the data which have been used for classification
will be shown.  Observations with a higher probability to be in the left
component are sown left, observations with a higher probability to be a member
of the second component are shown right.  On top (line) the estimated density
of the two components is shown.

```{r, fig = TRUE, fig.width = 10, fig.height = 6}
plot(mod, which = "hist")
```

Fitted probabilities (the probabilities obtained by the model based on
the training data set) can be accessed by calling `fitted`:

```{r}
# Class of fitted probabilities is a zoo object
class(fitted(mod))
summary(fitted(mod))
```

The method let's you access the fitted probabilities (`which = "probability"`),
a flag which comes with the fitted values (`which = "flag"`), or both
(`which = "both"`).

```{r}
# Probabilities only
tail(fitted(mod, which = "probability"))
# Flag only
tail(fitted(mod, which = "flag"))
# Both
tail(fitted(mod, which = "both"))
```

Explanation of the flag:
* `NA` if we have had missing data (missing covariates, or missing data
  to apply the `filter` rules provided when calling `foehnix`).
* `0` if observations have been outside the defined `filter`, not used
  for classification and thus the probability has been set to `0`.
* `1` if observations have been used for classification with the
  corresponding estimated probability.


## Predictions

A last important method is `predict`. The term predict refers to
statistical prediction, applying a statistical model to a data set
which has never been seen by the model (not included in the training data set).

As `foehnix` is a statistical classification method a prediction in this context
is not a forecast (like a weather forecast), but a classification to new data.
This can, for example, be used to retrieve foehn probabilities for new observation
in an operational context.

Assume we have estimated a `foehnix` mixture model and a set of new observations
from our two stations ([Ellboegen](reference/ellboegen.html) and [Sattelberg](reference/sattelberg.htlm))
have become available. We would like to get foehn probabilities this new data set.
Luckily we put aside 10 observations earlier (the `test` data set):


```{r}
# Calculate potential temperature difference (as we did above)
test$diff_t <- test$crest_t + 10.27 - test$t
print(test)
```

Given our model we can not perform the classification on this new data
set which has not been used for training the `foehnix` mixture model by calling:

```{r}
predict(mod, newdata = test)
```

For demonstration: we would have become the same (roughly the same, as
a slightly different training data set is used) if we would re-estimate
the model based on the whole data set (`train` and `test` combined)
and check the estimated probabilities of the last 10 entries, the very
same times we have put aside as our `test` data set:

```{r}
mod2 <- foehnix(diff_t ~ ff, data = rbind(train,test),
                filter = filter, switch = TRUE, verbose = FALSE)
print(tail(fitted(mod2, 1:2), 10))
```



















