```{r, echo = FALSE}
suppressPackageStartupMessages(library("foehnix"))
```

## Foehnix Demo

This page shows a _real world_ demo of the
automated foehn classification method **foehnix** based
on meteorological observations from two sites in Tyrol,
Austria. One station is located in the Brenner valley
south of Innsbruck on 1080 meters above mean sea level,
the second station is located close to the main alpine
ridge on 2107 meters above mean sea level
(see `?ellboegen`/`?sattelberg` for details).

```{r, echo = TRUE}
data("ellboegen"); data("sattelberg")
head(ellboegen);   head(sattelberg)
```

This example shows how to perform foehn classification
for Ellboegen using the information of the Sattelberg
station as additional source of information.
The Sattelberg station is located south of Ellboegen
close to the alpine crest and thus, during foehn events,
measuring the upstream air mass decending down on the
north side of the Alps.

### South Foehn Classification for Ellboegen

To perform the classification we first of all have to
prepare our data sets. `foehnix(...)` expects the input
to be of class `zoo` (zoo time series object; based on the
\emph{R} `zoo` package). Thus, after loading our example
data sets, we first have to convert the `data.frame`s into
valid `zoo` objects:

```{r}
# Loading observations (data.frame), convert to zoo
data("ellboegen", package = "foehnix")
data("sattelberg", package = "foehnix")
timefun    <- function(x) as.POSIXct(x,  origin = "1970-01-01", tz = "UTC")
ellboegen  <- zoo(ellboegen[,-1],  timefun(ellboegen$timestamp))
sattelberg <- zoo(sattelberg[,-1], timefun(sattelberg$timestamp))
```

As we want to use information from both stations: rename the
variables of the `sattelberg` object and combine the two
data sets into one `zoo` object called `data`.

```{r}
# Modify sattelberg variable names (crest_ identifies Sattelberg
# observations, our crest station) and combine both data sets.
names(sattelberg) <- paste0("crest_", names(sattelberg))
data <- merge(ellboegen, sattelberg)
```

In addition we are calculating the potential temperature difference
between the two sites using a dry adiabatic lapse rate of
one 10 degrees Celsius per 1000 meters. The difference in altitude
between the two stations is 1027 meters which yields:

```{r}
# Dry adiabatic temperature difference between 
# Sattelberg (data$crest_t) and Ellboegen (data$t) corrected by
# 1027/10 degrees.
data$diff_t <- data$crest_t + 10.27 - data$t
```

South foehn flow at the valley station in Ellboegen has a wind direction of
about 133 degrees. We are using a (relatively weak) wind direction filter with
a wind sector of +/-90 degrees around 133 degrees. For the classification only
observations/times are used where the corresponding wind direction (`dd`) lies
within the wind sector `>=43` and `<= 223` (180 degrees sector).
In addition we can make use of the information provided
by the crest station (Sattelberg) in several ways:

1. We use the temperature difference between the two
   stations as main variable for the classification (`diff_t`).
2. define wind sector filter at the target station (Ellboegen).
   Only use observations with a wind direction (`dd`) inside the
   wind sector of `>=43` and `<=223` degrees for classification
   (`dd = c(43, 223)`).
2. specify a second filter such that only observations are
   used for classification where the crest station shows
   a sutherly flow (wind direction within `>= 90` and `<= 270`,
   `sat_dd = c(90, 270)`).

```{r, echo = TRUE, results = 'hide'}
# Specify our wind filter rule(s)
filter <- list(dd = c(43, 223), crest_dd = c(90, 270))

# Estimate the mixture model for automated foehn classification
mod1 <- foehnix(diff_t ~ ff, data = data, filter = filter, switch = TRUE)
```

Let's check the estimated model:

```{r}
print(mod1)
print(coef(mod1))
```



