
```{r, echo = FALSE}
suppressPackageStartupMessages(library("foehnix"))
```

## Default Hovmöller Diagram

The generic function [`image`](reference/image.html) plots a Hovmöller diagram
of the estimated probabilities.

```{r image_default, echo = TRUE, figure = TRUE, fig.width = 12, fig.height = 7}
# Loading the demo data set
data <- demodata()

# Estimate the classification model
mod <- foehnix(diff_t ~ ff + rh, data = data,
               filter = list(dd = c(43, 223), crest_dd = c(90, 270)),
               switch = TRUE, verbose = FALSE)

# Create plot
image(mod)
```

By default, the Hovmöller diagram shows the frequency occurrence (estimated probability $\pi \ge 0.5$)
aggregated over 7 days (abscissa) on the same temporal resolution (ordinate) as the original data set.
The [`image`](reference/image.html) function allows for a high degree of customization. This page
contains some of the main features - details are provided on the corresponding
[manual page](references/image.html).



## Colors, Contours, and Limits

The plot can be customized by providing a vector of custom HEX colors (`col`), specify
custom limits on the abscissa (`xlim` $\in [0, 365]$), the ordinate (`ylim` $\in [0, 86400]$),
and for the range (`zlim`). In addition, `contours = TRUE` allows to add contour lines.
In addition, the two input arguments `deltat` and `deltad` can be used to change the aggregation intervals.

```{r image_colors_contours_limits, echo = TRUE, figure = TRUE, fig.width = 12, fig.height = 7}
image(mod,
      deltat = 3 * 3600,      # aggregation (time of the day): 3 hourly intervals
      deltad = 14,            # aggregation (day of the year): two-week intervals
      col = colorspace::sequential_hcl(21, "BuPu", rev = TRUE), # custom colors
      contours = TRUE,        # enable contour lines
      contour.col = "gray20", # custom contour colors
      main = "Custom Colors and Contours")
```

The limits can be used to adjust the area to be plotted.

```{r image_custom_limits, echo = TRUE, figure = TRUE, fig.width = 12, fig.height = 7}
image(mod,
      col = colorspace::sequential_hcl(21, "BuPu", rev = TRUE),
      contours = TRUE,
      contour.col = "white",
      xlim = c(91, 151),
      ylim = c(9, 18) * 3600,
      zlim = c(0, 0.5),
      main = "Custom Plot and Color Limits",
      xlab = "April and May",
      ylab = "09:00 UTC - 18:00 UTC")

```


This is especially useful if one
has a data set which only covers a short time period (e.g., only several months). **Please note**
that, for demonstration purposes, we will use a subset of the [`demodata`](references/demodata.html)
data set which comes with an hourly resolution. Thus, the number of observations to train the
`foehnix` models is rather short and may not lead to robust estimates. In a real-world application
you may use a higher temporal resolution (e.g., 10min observations).


```{r image_custom_data1, echo = TRUE, figure = TRUE, fig.width = 10, fig.height = 5}
# Subsetting the demo data set
# - data_subset: take data from September 2015 trough April 2016
data_subset <- window(data, start = as.POSIXct("2015-09-01 01:00"), end = as.POSIXlt("2016-04-30"))

# Estimate foehnix classification models
mod_subset <- foehnix(diff_t ~ ff + rh, data = data_subset,
                      filter = list(dd = c(43, 223), crest_dd = c(90, 270)),
                      switch = TRUE, verbose = FALSE)

# Plotting "mod1" with custom x-limits
xlim <- as.POSIXlt(range(index(data_subset)))$yday
image(mod_subset, xlim = xlim,
      main = sprintf("Custom x limits from %d - %d (Julian day)", xlim[1L], xlim[2L]))
```

**Note** that our data subset contains observations from September 2015 trough April 2016
(one winter; over new years eve). If decreasing x-limits are provided the
[`image`](references/image.html) function adjusts itself to show one continuous period
over new years eve (here: autumn over winter to spring).



## Custom Aggregation

The argument `FUN` can either be a custom _R_ function or one of the following:

* `"freq"`: frequency of occurrence ($\frac{\#\{\pi \gt 0.5\}}{N}$) plotting frequencies.
* `"mean"`: plots the mean probability ($\frac{1}{N} \sum_{i}^{N} \pi_i$).
* `"occ"`: occurrence of foehn (absolute number; $\#\{pi_i \gt 0.5\}$).
* `"noocc"`: the inverse to `"occ"` (absolute number; $\#\{pi_i < 0.5\}$).

Some examples:

```{r}
image(mod, FUN = "occ", col = rainbow(10), main = "foo")
```
