
```{r, echo = FALSE}
suppressPackageStartupMessages(library("foehnix"))
```

This page shows how one can import observational data in such a way that
they can be used with _foehnix_. _foehnix_ uses [`zoo`](https://cran.r-project.org/package=zoo)
objects (time series objects) to handle the input data.
This page is not intended to be a complete manual or introduction to 
[zoo](https://cran.r-project.org/package=zoo)!
The [zoo](https://cran.r-project.org/package=zoo) itself comes with a range
of vignettes including commonly asked questions and detailed introductory material.
If you are not familiar with the package and/or the following examples are not
enough to get started, please visit the
[zoo](https://cran.r-project.org/package=zoo) package website to get more information.


* [cran.r-project.org/package=zoo](https://cran.r-project.org/package=zoo)



## Example A


Imagine you have a [CSV](https://en.wikipedia.org/wiki/Comma-separated_values) file which looks as follows:

```{r data set A show, echo = FALSE, result = "asis"}
# Read the first few lines of the demo data set. Path relative to the
# 'vignettes' folder.  Note: new files have to be copied into the docs folder
# using pkgdown::init_site().
x <- readLines("../pkgdown/data/ellboegen_A.csv", n = 10)
cat(paste(x, collapse = "\n"), "\n")
```

The output shows the first 10 rows (header and 9 data rows) of the demo data set
"[ellboegen_A.csv](../pkgdown/data/ellboegen_A.csv)" with hourly observations.
This is the "default _R_ CSV file format" (see `write.csv`) where the values
are comma-separated, strings are quoted, missing values are simply "" (no character),
and no unnecessary blanks are added to the file. The colums conntain:

* `date_time`: a column with date and time information (default english date format)
* `dd`: wind direction in degrees
* `ff`: wind speed in meters per second
* `rh`: relative humidity
* `t`: 2m air temperature

### Import data

To import the data set and convert the data into `zoo`, all we have to do is to
use the function `zoo::read.zoo` like this:

```{r data set A import}
# Load package 'zoo'
library("zoo")
# Import the data set
data <- read.zoo("../pkgdown/data/ellboegen_A.csv", format = "%Y-%m-%d %H:%M:%S",
                 tz = "UTC", sep = ",", header = TRUE)
```

The call `read.zoo(...)` loads the demo data set (here `../pkgdown/data/ellboegen_A.csv`),
specifies the date/time format (`format`; `read.zoo` expects this information in the first column by default),
a time zone argument (`tz`), the separator in the CSV file (`sep`), and that we do have
a header line (`header = TRUE`).
`read.zoo` calls `read.table`, tries to extract the date/time information, and creates a `zoo` object.

```{r}
head(data, n = 4)
```

### Import and combine data

In case a crest station is available, we need to combine observations from two stations,
in this case two demo data sets for Ellbögen ([ellboegen_A.csv](../pkgdown/data/ellboegen_A.csv)) and Sattelberg
([sattelberg_A.csv](../pkgdown/data/sattelberg_A.csv); our crest station).
Both files are in the very same format. Thus, we can load both data sets as follows:


```{r data set A import both}
# Load package 'zoo' if not already loaded
library("zoo")
# Import data set 'Ellboegen'
ell <- read.zoo("../pkgdown/data/ellboegen_A.csv", format = "%Y-%m-%d %H:%M:%S",
                tz = "UTC", sep = ",", header = TRUE)
# Import data set 'Sattelberg'
sat <- read.zoo("../pkgdown/data/sattelberg_A.csv", format = "%Y-%m-%d %H:%M:%S",
                tz = "UTC", sep = ",", header = TRUE)
```

All we have to do is to combine the two objects `ell` and `sat` which an be done
using the function `zoo::merge`. `zoo::merge` automatically takes care that the two
time series are matched propperly (conditional on date/time).

```{r data set A combine 1}
# Combine data
data <- merge(ell, sat)
head(data, n = 8)
```

By default, missing data are filled with `NA` (missing value). As the demo data set for
station Sattelberg starts four hours later than the one for Ellbögen, the first for rows
for `sat` (`01:00:00` to `04:00:00`) are empty.
As the variables in both files are the very same, _R_ automatically adds `.ell` or `.sat`
to the original variable names (columns in the CSV file). In case we would like to have nicer
names, we could prepare them manually, e.g.,:

```{r data set A rename}
# Rename the variables in 'sat'
names(sat) <- paste("crest", names(sat), sep = "_")
# Show new names
names(sat)
```

And combine the data set once again (overwrites `data`):

```{r data set A combine 2}
# Combine (again)
data <- merge(ell, sat)
head(data, n = 8)
```

And that's it. This object (`data`) could now be used as input for the
`foehnix` method.




## Example B

The next demo data contains the very same as the data set above, however, the
are distinct differences in the format of the CSV file
(see [ellboegen_B.csv](../pkgdown/data/ellboegen_B.csv), [sattelberg_B.csv](../pkgdown/data/sattelberg_B.csv)):

```{r data set B show, echo = FALSE, result = "asis"}
# Read the first few lines of the demo data set. Path relative to the
# 'vignettes' folder.  Note: new files have to be copied into the docs folder
# using pkgdown::init_site().
x <- readLines("../pkgdown/data/ellboegen_B.csv", n = 10)
cat(paste(x, collapse = "\n"), "\n")
```

In contrast to 'data set A' the file solely contains numeric values - except the missing
values (`missing`), the date/time information is coded as integer (`YYYYmmddHHMMSS`; last column)
and there is no explicit column separator (columns are separated by one or multiple blanks).

### Import data

To be able to import the data set we do have to specify the format. In contrast to
'data set A' we need:

* different `format`
* a custom function `FUN` to convert the integers (column `date_time`) into `POSIXt`
* an additional argument `index.column = "date_time"` to tell `zoo` where the date/time information is stored
* an input `na.strings` which defines how the "missing values" in the CSV file look like

Overall we can read the file(s) like this:

```{r}
# Load library (if not yet done)
library("zoo")
# Custom function to convert the integers (data_time) into POSIXct
FUN = function(x, format, tz, ...) as.POSIXct(strptime(sprintf("%.0f", x), format), tz = tz)
# Import data set
data <- read.zoo("../pkgdown/data/ellboegen_B.csv", format = "%Y%m%d%H%M%S", tz = "UTC",
                 FUN = FUN, index.column = "date_time", 
                 header = TRUE, na.strings = "missing")
head(data, n = 3)
```

### Import and combine

We can do the very same for the second data set (Sattelberg) and combine the the data from
Sattelberg and Ellbögen:

```{r data set C combine}
# Loading library 'zoo'
library("zoo")
# User-defined function to convert date/time information
FUN = function(x, format, tz, ...) as.POSIXct(strptime(sprintf("%.0f", x), format), tz = tz)
# Read ellboegen data set
ell <- read.zoo("../pkgdown/data/ellboegen_B.csv", format = "%Y%m%d%H%M%S", tz = "UTC",
                FUN = FUN, index.column = "date_time", 
                header = TRUE, na.strings = "missing")
sat <- read.zoo("../pkgdown/data/sattelberg_B.csv", format = "%Y%m%d%H%M%S", tz = "UTC",
                FUN = FUN, index.column = "date_time", 
                header = TRUE, na.strings = "missing")
# Rename columns in 'sat'
names(sat) <- paste("crest", names(sat), sep = "_")
# Combine
data <- merge(ell, sat)
# Show first 8 entries
head(data, n = 8)
```

# Other Formats

The method `zoo::read.zoo` has a wide range of arguments (see also `?read.table`)
which allows to import a wide range of possible formats. If you need more information
please visit the [zoo](https://cran.r-project.org/package=zoo) package information page
on [CRAN](https://cran.r-project.org) where you can find manuals and vignettes with more
details about the [zoo](https://cran.r-project.org/package=zoo) package and how to import/create
`zoo` time series objects in _R_.












