```{r, echo = FALSE}
suppressPackageStartupMessages(library("foehnix"))
```

## Foehnix Demo

`foehnix` objects also provide asymptotic inference of the estimated
coefficients of the mixture model (see [statistical model](mixedmodel.html) and
[logistic regression with IWLS](logisticregression.html)).

Let's start with estimating our demo `flexmix` foehn diagnosis model. We have
already prepared the data sets (object `data`). Details how to generate
the data object and more information about the `foehnix` model specification
can be found [on the getting started](foehnix.html) manual page.

```{r, echo = FALSE}
# Loading combined demo data set
data <- demodata()
```

```{r, echo = TRUE, results = "hide"}
# Estimate the model
mod <- foehnix(diff_t ~ ff + rh, data = data)
```

The `summary` method prints the test statistics for both
parts of the mixture model, the component model and the
concomitant model (if specified).

```{r, echo = TRUE}
summary(mod, detailed = TRUE)
```

## Component Model Inference

The inference for the two location parameters $\mu_1$ and $\mu_2$ of the two
components are based on the asymptotic theory.  We expect that the estimates of
our coefficients are unbiased. Thus, the expectation of our estimated
coefficients is the estimated coefficient itself
($\text{E}(\hat{\mu}_1) = \mu_1$, $\text{E}(\hat{\mu}_2) = \mu_2$).

In a general form the covariance matrix of a liner model (one component,
unweighted) for a set of $i = 1, \dots, N$ observations can be expressed as follows:

* $\text{Cov}(\beta) = \frac{1}{N - P} \mathit{\epsilon}^\top \mathit{\epsilon} (\mathbf{X}^\top \mathbf{X})^{-1}$

... where $N$ is the sample size, $P$ the number of parameters or covariates,
$\mathit{\epsilon} = \mathit{y} - \mu_\bullet$ the model residuals, and $\mathbf{X}$ the model matrix of
the linear model.  In case of a `foehnix` model we have two components where
each component consists of an intercept only model ($\mu_1$ and $\mu_2$ do not
depend on additional covariates). Thus, $P = 1$ and $\mathbf{X}$ is an $N
\times 1$ matrix with 1s. The estimates are based on a set of weighted
observation $y$, in this example `diff_t` ($y =$ `diff_t`).
The weights are the a-posteriori probabilities $\hat{\mathit{p}}$ (the foehn probabilities) of
the `foehnix` model and have to be taken into account when calculating the standard errors.
With these weights the standard error for _component 2_ can be written as:

* $\hat{\text{sd}}(\mu_2)^2 = \frac{1}{\sum_{i=1}^N \hat{p}_i - 1} (\mathit{\epsilon} \cdot \hat{\mathit{p}})^\top (\mathit{\epsilon} \cdot \hat{\mathit{p}}) \big((\mathbf{X} \cdot \hat{\mathit{p}})^\top (\mathbf{X} \cdot \hat{\mathit{p}})\big)^{-1}$

... or much simpler:

* $\hat{\text{sd}}(\mu_2)^2 = \frac{1}{\sum_{i=1}^N \hat{p}_i^2 \cdot (\sum_{i=1}^N \hat{p}_i - 1)} \sum_{i=1}^N \big((y_i - \mu_2) \cdot \hat{p}_i\big)^2$

The same holds for _component 1_ except that our weights are $1 - \hat{\mathit{p}}$:

* $\hat{\text{sd}}(\mu_1)^2 = \frac{1}{\sum_{i=1}^N (1 - \hat{p}_i)^2 \cdot (\sum_{i=1}^N (1 - \hat{p}_i) - 1)} \sum_{i=1}^N \big((y_i - \mu_2) \cdot (1 - \hat{p}_i)\big)^2$

## Concomitant Model Inference

If a concomitant model has been specified `summary` will also return the
corresponding z statistics for the estimated regression coefficients of the
logistic regression model (see [logistic regression with IWLS](logisticregression.html)).

The covariance matrix of a logistic logistic regression model with the regression
coefficients $\mathit{\alpha}$ (with $P$ coefficients) with a dispersion parameter of 1
of the binomial family is given as:

* $\hat{\text{cov}}(\mathit{\alpha}) = (\mathbf{X} \mathit{\omega})^\top (\mathbf{X} \mathit{\omega})$

where $\mathbf{X}$ is the model matrix of the concomitant model (un-standardized) of
dimension $N \times P$ and $\mathit{\omega} = (\mathit{\pi} * (1 - \mathit{\pi}))^\frac{1}{2}$.
$\pi$ is final response, probability returned by the logistic regression model.
The diagonal of the covariance matrix contains the variances of the estimated
regression coefficients. Thus,

* $\hat{\text{sd}}(\mathit{\alpha}) = \Big( \text{diag}\big((\mathbf{X} \mathit{\omega})^\top (\mathbf{X} \mathit{\omega}) \big) \Big)^\frac{1}{2}$.









