
## Simulation

The `foehnix` package allows to estimate Gaussian and logistic
mixture models with and without censoring and truncation.
The package default is _Gaussian_ without censoring or truncation
which typically works quite well. Fore those interested in
censoring and truncation please have a look to the **Advanced**
pages of this manual.

This page shows two examples using simulated data from a Gaussian
mixed distribution and a logistic mixed distribution without
additional concomitants and the corresponding model fits.


```{r, echo = FALSE, fig = FALSE}
suppressPackageStartupMessages(library("zoo"))
suppressPackageStartupMessages(library("foehnix"))
```

### Gaussian

Simulate random data from a mixed Gaussian distribution
using the following _true parameters_:

```{r, fig = FALSE}
true <- list(mu = c(1, 5), sigma = c(1, 2))
```

The first component has a mean (mu) of `1.0` and a standard
deviation (sigma) of `1.0`, the second component a mean of
`5.0` and a corresponding standard deviation of `2.0`.
Given these parameters we can draw a set of random
numbers from a two-component Gaussian mixture distribution:

```{r, fig = FALSE}
# Drawing 15000 random numbers from Gaussian mixed distribution
family <- foehnix:::foehnix_gaussian();   set.seed(666)
data <- data.frame(y = family$r(c(10000, 5000), mu = true$mu, sigma = true$sigma))
# Convert data to zoo (time series object) as foehnix expects
# time series input data
data <- zoo(data, as.Date(1:nrow(data), origin = "1970-01-01"))
```

Our `data` set now consists of `15000` random numbers from the two
components specified above (`true`) whereof two third come from the
first component, one third from the second component. Or in other words:
the _true probability_ that an observation comes from the second 
component is `0.333`.
We can now _estimate a foehnix model_:

```{r, fig = FALSE, results = 'hide'}
# Estimate foehnix model
mod <- foehnix(y ~ 1, data = data)
```
```{r, fig = FALSE}
coef(mod)
```

By default, a Gaussian mixture model is estimated. `coef(mod)` is used
to extract the estimated coefficients:

* Estimated parameters for component 1:
    * `mu` (or mean): `r round(coef(mod)["mu1"],2)` (true was `r true$mu[1L]`)
    * `sigma` (or standard deviation): `r round(coef(mod)["sigma1"],2)` (true was `r true$sigma[1L]`)
* Estimated parameters for component 2:
    * `mu` (or mean): `r round(coef(mod)["mu2"],2)` (true was `r true$mu[2L]`)
    * `sigma` (or standard deviation): `r round(coef(mod)["sigma2"],2)` (true was `r true$sigma[2L]`)


```{r, fig.width = 8, fig.height = 6}
# Path of log-likelihood sum during optimization
plot(mod, which = "loglik")

# Plotting mixed distribution
# Calculate density of the two components
x <- seq(min(data$y), max(data$y), length = 501)
density1 <- family$d(x, coef(mod)["mu1"], coef(mod)["sigma1"])
density2 <- family$d(x, coef(mod)["mu2"], coef(mod)["sigma2"])
hist(data$y, breaks = 100, freq = FALSE,
     xlab = "simulated data",
     main = "Histogram of Simulated Data\nPlus Estimated Mixed Model Density")
lines(x, density1 * (1 - mean(fitted(mod))), col = 2, lwd = 2)
lines(x, density2 * mean(fitted(mod)),       col = 4, lwd = 2)
lines(x, density1 * (1 - mean(fitted(mod))) + density2 * mean(fitted(mod)), col = 6, lty = 3, lwd = 3)
legend("topright", col = c(2, 4, 6), lwd = 2, lty = c(1, 1, 3), bty = "n",
       legend = c("Component 1", "Component 2", "Mixture Density"))
```




### Logistic

The very same can be done using a two-component logistic mixture model.
As above a set of `15000` random values from the mixture model are drawn
and modelled using foehnix with the same _true parameters_ as above for the
Gaussian model.

```{r, fig = FALSE}
# The true parameters of the distribution
true <- list(mu = c(1, 5), sigma = c(1, 2))

# Simulate data based on logistic distribution and create time series
family <- foehnix:::foehnix_logistic();    set.seed(666)
data <- data.frame(y = family$r(c(10000, 5000), mu = true$mu, sigma = true$sigma))
data <- zoo(data, as.Date(1:nrow(data), origin = "1970-01-01"))
```

All we have to do is to specify `family = "logistic"` when calling the
`foehnix(...)` method:


```{r, fig = FALSE, results = 'hide'}
# Estimate foehnix model
mod <- foehnix(y ~ 1, data = data, family = "logistic")
```
```{r, fig = FALSE}
coef(mod)
```

* Estimated parameters for component 1:
    * `mu` (or mean): `r round(coef(mod)["mu1"],2)` (true was `r true$mu[1L]`)
    * `sigma` (or standard deviation): `r round(coef(mod)["sigma1"],2)` (true was `r true$sigma[1L]`)
* Estimated parameters for component 2:
    * `mu` (or mean): `r round(coef(mod)["mu2"],2)` (true was `r true$mu[2L]`)
    * `sigma` (or standard deviation): `r round(coef(mod)["sigma2"],2)` (true was `r true$sigma[2L]`)

```{r, fig.width = 8, fig.height = 6}
# Path of log-likelihood sum during optimization
plot(mod, which = "loglik")

# Mean probability
meanprob <- mean(fitted(mod), na.rm = TRUE)

# Plotting mixed distribution
# Calculate density of the two components
x <- seq(min(data$y), max(data$y), length = 501)
density1 <- family$d(x, coef(mod)["mu1"], coef(mod)["sigma1"])
density2 <- family$d(x, coef(mod)["mu2"], coef(mod)["sigma2"])
hist(data$y, breaks = 100, freq = FALSE,
     xlab = "simulated data",
     main = "Histogram of Simulated Data\nPlus Estimated Mixed Model Density")
lines(x, density1 * (1 - mean(fitted(mod))), col = 2, lwd = 2)
lines(x, density2 * mean(fitted(mod)),       col = 4, lwd = 2)
lines(x, density1 * (1 - mean(fitted(mod))) + density2 * mean(fitted(mod)), col = 6, lty = 3, lwd = 3)
legend("topright", col = c(2, 4, 6), lwd = 2, lty = c(1, 1, 3), bty = "n",
       legend = c("Component 1", "Component 2", "Mixture Density"))
```

