
```{r, echo = FALSE}
suppressPackageStartupMessages(library("foehnix"))
```

## Time Series Plot

TODO: Explain the default plot.

```{r tsplot default, echo = TRUE, figure = TRUE, fig.width = 15, fig.height = 9}
# Loading the demo data set for Tyrol
data <- demodata("tyrol")
# Faking ffx (wind gusts) as we don't have them in the data set.
data$ffx       <- runif(nrow(data), 1.4, 1.7) * data$ff
data$crest_ffx <- runif(nrow(data), 1.5, 2.5) * data$crest_ff
# Show data set 
head(data, n = 3)

# Estimate a foehnix classification model
filter <- list(dd = c(43, 223), crest_dd = c(90, 270))
mod <- foehnix(diff_t ~ ff + rh, data = data, filter = filter,
               switch = TRUE, verbose = FALSE)

# Alternative model. Does not use data from the crest station Sattelberg
alt_filter <- list(dd = c(43, 223))
alt_mod <- foehnix(ff ~ rh, data = data, filter = alt_filter, verbose = FALSE)


# Plotting windroses
tsplot(list("Enhanced" = mod, "Alternative (basic)" = alt_mod), start = "2015-05-01", end = "2015-05-10")
```

## Different Styles

TODO: The `style` argument (input to `tsplot.control`) allows to change between
different styles provided by the foehnix package. The styles can be modified
by the user (see next section).

```{r tsplot advanced, echo = TRUE, figure = TRUE, fig.width = 15, fig.height = 9}
tsplot(list("Enhanced" = mod, "Alternative (basic)" = alt_mod),
       start = "2015-05-01", end = "2015-05-10",
       style = "advanced")
```

```{r tsplot bw, echo = TRUE, figure = TRUE, fig.width = 15, fig.height = 9}
tsplot(list("Enhanced" = mod, "Alternative (basic)" = alt_mod),
       start = "2015-05-01", end = "2015-05-10",
       style = "bw")
```

## Modify styles

TODO: No matter which style one takes, one is always allowd to overwrite
specific properties of such as line type, color, ....
This allows for high customization of the plots.

As an example we change the appearance of the observed air temperature (`t`)
and disable two parameters (`crest_t` and `diff_t`) by setting them to `NULL`.

```{r tsplot customized advanced, echo = TRUE, figure = TRUE, fig.width = 15, fig.height = 9}
tsplot(list("Enhanced" = mod, "Alternative (basic)" = alt_mod),
       start = "2015-05-01", end = "2015-05-10",
       t       = list(col = "blue", lty = 3, lwd = 4),
       crest_t = NULL, # Disable crest temperature
       diff_t  = NULL, # Disable temperature difference
       style = "advanced")
```

The same can, of course, be done with the black and white style:

```{r tsplot customized bw, echo = TRUE, figure = TRUE, fig.width = 15, fig.height = 9}
tsplot(list("Enhanced" = mod, "Alternative (basic)" = alt_mod),
       start = "2015-05-01", end = "2015-05-10",
       t = list(col = "orange", type = "o", pch = 18),
       style = "bw")
```

## Highlight Wind Sectors

TODO: Wind sectors (as for the windrose plot) can be used to set specific highlights.
This optional argument is also forwarded to `tsplot.control`.

```{r tsplot windsector, echo = TRUE, figure = TRUE, fig.width = 15, fig.height = 9}
tsplot(list("Enhanced" = mod, "Alternative (basic)" = alt_mod),
       start = "2015-05-01", end = "2015-05-10",
       windsector = list("Sector A" = c(88, 178), "Test Sector" = c(340, 20)),
       foehnix_windsector = list(col = "orange"), # Custom color
       style = "bw")
```

