
```{r, echo = FALSE}
suppressPackageStartupMessages(library("foehnix"))
```

# Data Set Description

The "Californian" data set consists of hourly meteorological observations
from station "Viejas Casino and Ranch" and station "Lucky Five Ranch"
located in South California. 

Viejas is located at the foot of the westerly slope of the Sierra Nevada
mountain range and exhibits strong easterly winds during downslope wind
situations. The Lucky Five Ranch is located northeast of and provides
information about the upstream air mass for the classification algorithm.


```{r leaflet, echo = FALSE, fig.width = 10, fig.height = 5}
library("leaflet")
library("sp")

stations <- data.frame(lon  = c(-116.70437, -116.528),
                       lat  = c(32.84559, 32.9331),
                       alt  = c(715, 1445),
                       name = c("Viejas", "Lucky Five Ranch"),
                       stringsAsFactors = FALSE)

b <- list(x0 = min(stations$lon) - .2, x1 = max(stations$lon) + .2,
          y0 = max(stations$lat) + .2, y1 = max(stations$lat) + .2)

m <- leaflet(width = "100%") %>% addTiles() %>% fitBounds(b$x0, b$y0, b$x1, b$x2)
m <- setView(m, mean(stations$lon), mean(stations$lat), zoom = 10.5)
for (i in 1:nrow(stations)) {
    m <- addPopups(m, stations$lon[i], stations$lat[i],
                   sprintf("Station %s, %d m a.m.s.l.", stations$name[i],
                           stations$alt[i]))
}
m <- addProviderTiles(m, "OpenTopoMap")
m
```


# Loading the Data Set

[`demodata("california")`](references/demodata) returns a data set which
combines hourly meteorological observations of both sites (Viejas; Lucky Five).
In addition, the potential temperature difference between the two stations
is calculated by reducing the dry air temperature from "Lucky Five Ranch"
to the height of "Viejas" (dry adiabatic lapse rate of 1K per 100m;
stored on `diff_temp`).
For details see [`demodata`](references/demodata.html).

```{r load_california}
data <- demodata("california")
head(data, n = 3)
# Check if our object is a numeric zoo object:
c("is.zoo" = is.zoo(data), "is.numeric" = is.numeric(data))
```

Missing values in the
data set (`NA`) are allowed and will be properly handled by all functions.
One restriction is that the time series object has to be regular (but not
strictly regular). "Regular" means that the time steps have to be divisible
by the smallest time step, "strictly regular" means that we have no missing
observations (if our smallest time interval is 1 hour observations have
to be available _every hour_ have to be available to be strictly regular).
The [`foehnix`](references/foehnix.html) will inflate the data set and make
it strictly regular, if needed.

```{r check_regular}
c("is regular" = is.regular(data),
  "is strictly regular" = is.regular(data, strict = TRUE))
```



After preparing the data set (regular or strictly regular `zoo   object with
`numeric` values) we can investigate the observed wind information.

```{r emp_windrose, fig = TRUE, fig.width = 12, fig.height = 6}
par(mfrow = c(1,2))
# Observed wind speed/wind direction "Viejas"
windrose(data, ddvar = "wind_direction", ffvar = "wind_speed",
         main = "Windrose\nViejas Casino and Resort",
         breaks = seq(0, 22, by = 2))
# Observed wind speed/wind direction "Lucky Five"
windrose(data, ddvar = "crest_wind_direction", ffvar = "crest_wind_speed",
         main = "Windrose\nLucky Five Ranch", hue = 270,
         breaks = seq(0, 22, by = 2))
```

Given the plots above we define the `foehn` wind direction
at Viejas between 305 and 160 degrees (a `r 360 - (305 - 160)` degree
wind sector centered northeast). This wind sector can be chosen
rather wide, but should leave out non-foehn wind directions to exclude
upslope winds. The wind sector(s) can be added to the
[`windrose`](references/windrose.html) as follows:

```{r, fig = TRUE, fig.width = 8, fig.height = 8}
# Windrose plot with custom variable names (ddvar, ffvar),
# title, breaks, polygon borders, and the wind sector from above
# with a custom color.
windrose(data, ddvar = "wind_direction", ffvar = "wind_speed",
         main = "Windrose Viejas with Custom foehn Wind Sector",
         breaks = seq(0, 16, by = 2),
         windsector = list(c(305, 160)),
         windsector.col = "#DFEFF6",
         border = "gray50", lwd = .5)
```

The `windsector`
is solely used for visual justification, the same restriction will
be used in the following step when estimating the [`foehnix`](references/foehnix.html)
classification model.


# Estimate Classification Model

The next step (the core feature of this package) is to estimate
the two-component mixture model for foehn classification.
The following model assumptions are used here:

* **Main variable**: `diff_temp` (potential temperature difference) is used as
  the main covariate to separate 'foehn' from 'no foehn' events.
* **Concomitant variable**: `wind_speed` (wind speed at target station Viejas).
* **Wind filter**: the `wind_direction` at station Viejas has to lie within
    305 and 160 degrees (northeasterly wind direction; see above).
* **Option switch:** `switch = TRUE` as high `diff_temp` indicate stable stratification (no foehn).

```{r estimate_model, results = "hide"}
# Estimate the foehnix classification model
mod <- foehnix(diff_temp ~ wind_speed,
               data   = data,
               switch = TRUE,
               filter = list(wind_direction = c(305, 160)))
```

### Model Summary

```{r model_summary, echo = FALSE}
# Model summary
summary(mod, detailed = TRUE)
```

The full data set contains $N = `r nrow(mod$data)`$ rows, $`r nrow(data)`$ from
the data set itself (`data`) and $`r mod$inflated`$ due to inflation used to make the
time series object strictly regular.

Thereof, $`r length(mod$filter_obj$ugly)`$ are not considered
due to missing data, $`r length(mod$filter_obj$bad)`$ as they do not fulfil the filter constraint
(`wind_direction` outside defined wind sector), wherefore the final
model is based on $`r length(mod$filter_obj$good)`$ observations (or rows).

One good indication whether the model well separates the two clusters
is the "_Cluster separation_" output ([`summary`](references/summary.html))
or the posterior probability plot:

```{r}
# Cluster separation (summary)
summary(mod)$separation
```

The `separation` matrix shows the prior probabilities,
the size (number of observations assigned to each component; posterior probability),
number of probabilities exceeding a threshold (default `eps = 1e-4`), and the
ratio between the latter two. Ratios close to $1.0$ indicate that the two clusters
are well separated ($ratio > 0.5$ are already good for this application).

Another indication is the `which = "posterior"` plot which shows the empirical
histogram of estimated probabilities (for within-windsector observations).
Point masses around $0.0$ and $1.0$ indicate that we have two well separated
clusters (the probability to fall in one of the clusters is always close to
either $0$ or $1$).

```{r, fig = TRUE, fig.width = 8, figh.height = 4}
plot(mod, which = "posterior", breaks = seq(0, 1, by = 0.05)) 
```




### Model coefficients

The following parameters are estimated for the two `r mod$control$family$name` clusters:

* No-foehn cluster: $\mu_1 = `r round(coef(mod)["mu1"], 2)`$,
                    $\sigma_1 = `r round(coef(mod)["sigma1"], 2)`$ (parameter scale)
* Foehn cluster: $\mu_2 = `r round(coef(mod)["mu2"], 2)`$,
                    $\sigma_2 = `r round(coef(mod)["sigma2"], 2)`$ (parameter scale)
* Concomitant model: `r ifelse(sign(coef(mod)["wind_speed"]), "positive", "negative")`
    `wind_speed` effect, `r sprintf("%+.1f", 100 * exp(coef(mod)["wind_speed"]) - 100)` percent per
    $m~s^{-1}$

```{r model_coef}
coef(mod)
```


<div class="hide-Rinput">

```{r plot_mod_2dimg, fig = TRUE, fig.width = 8, fig.height = 6}
# Grid definition for 2-D plot
tmp <- list(diff_temp      = seq(-5, +12, by = .5),
            wind_speed     = seq(0, 12, by = .5),
            wind_direction = 45)
# Expand grid to create 'newdata' data.frame
grd  <- do.call(expand.grid, tmp)
mat  <- matrix(grd$diff_temp, nrow = length(tmp$diff_temp))
# Predict probabilities
pred <- matrix(predict(mod, newdata = grd)$prob, nrow = length(tmp$diff_temp))
# Create plot
cols <- colorspace::diverging_hcl(51, p1 = 5)
par(mar = rep(.5, 4), oma = c(3.1, 3.1, 2.1, 3.1))
layout(matrix(1:2, nrow = 1L), widths = c(10, 1))
image(x = tmp$diff_temp, y = tmp$wind_speed, z = pred, col = cols, zlim = c(0,1))
mtext(side = 1, line = 2.5, "diff_temp")
mtext(side = 2, line = 2.5, "wind_speed")
image(matrix(seq(0, 1, length = 51), nrow = 1),
      col = cols, yaxt = "n", xaxt = "n")
mtext(side = 4, line = 2.5, "probability")
mtext(side = 3, outer = TRUE, font = 2, cex = 1,
      "Predicted Probability | Wind Direction = 45")
axis(side = 4, at = seq(0, 1, by = .1))
```

</div>

<div class="hide-Rinput">

```{r plot_mod_2dpoints, fig = TRUE, fig.width = 8, fig.height = 6}
# Grid definition for 2-D plot
prob <- fitted(mod)[mod$filter_obj$good] # Estimated probabilities
cols <- cut(prob, breaks = seq(0, 1, by = 0.1), include.lowest = TRUE,
            labels = colorspace::diverging_hcl(10))
# Prepare temporary data set (out of those used for classification)
tmp  <- as.data.frame(subset(mod$data[mod$filter_obj$good, ],
                             select = c(wind_speed, diff_temp)))
tmp$col <- cols      # Add color information
tmp <- unique(tmp)   # We only need unique entries
# Create plot
with(tmp,
     plot(wind_speed ~ diff_temp, col = as.character(col), pch = 19,
          main = "Fitted Probabilities | within Wind Sector"))
legend("topright", fill = colorspace::diverging_hcl(10), bty = "n",
       title = "probability",
       legend = levels(cut(0, breaks = seq(0, 1, by = 0.1), include.lowest = TRUE)))
```

</div>


# Graphical Model Assessment

A [`foehnix`](references/foehnix.html) object comes with generic plots for graphical model
assessment.

The following figure shows the 'log-likelihood contribution' of

* the main **component** (left hand side of formula),
* the **concomitant** model (right hand side of formula),
* and the **full** log-likelihood sum which is maximised by
  the optimization algorithm.

The abscissa shows (by default) the logarithm of the iterations during 
optimization. 

```{r plot_loglikcontribution, fig = TRUE, fig.width = 12, fig.height = 6}
# Log-likelihood contribution
plot(mod, which = "loglikcontribution")
```

In addition, the coefficient paths during optimization can be visualized:

```{r plot_coefpath, fig = TRUE, fig.width = 12, fig.height = 6}
# Coefficient path
plot(mod, which = 3L)
```

The left plot shows the parameters of the two components
($\mu_1$, $\log(\sigma_1)$, $\mu_2$, $\log(\sigma_2)$), the
right one the standardized coefficients of the concomitant model.

Last but not least a histogram with the two clusters is plotted.
`which = "hist"`creates an empirical density histogram separating "no foehn"
and "foehn" events adding the estimated distribution for these two clusters.

```{r plot_histogram, fig = TRUE, fig.width = 8, fig.height = 3.5}
devtools::load_all("..")
plot(mod, which = "hist")
```



# Time Series Plot

The Californian demo data set has non-standard variable names (by purpose).
Thus, when calling `tsplot` (time series plot) we do have to manually specify
these names.

```{r tsplot, fig = TRUE, fig.width = 12, fig.height = 15}
# Some smaller quality issues in the data (should not be a big deal)
start <- as.POSIXct("2012-03-01")
end   <- as.POSIXct("2012-03-12")

# As we dont have the standard names: re-specify variable names.
# In addition, use 'style = "advanced"' to show more details.
tsplot(mod, style = "advanced", 
       diff_t = "diff_temp",     rh = "relative_humidity",
       t   = "air_temp",         crest_t   = "crest_air_temp",
       dd  = "wind_direction",   crest_dd  = "crest_wind_direction",
       ff  = "wind_speed",       crest_ff  = "crest_wind_speed",
       ffx = "wind_gust",        crest_ffx = "crest_wind_gust",
       windsector = list(wind_direction = c(305, 160)),
       start = start, end = end)
```


# Wind Rose Plot

```{r windrose, fig = TRUE, fig.width = 12, fig.height = 5}
devtools::load_all("..")
windrose(mod, dd = "wind_direction", ff = "wind_speed",
         type = "hist", which = c("foehn", "nofoehn"),
         windsector = list(c(305, 160)),
         breaks = seq(0, 16, by = 2))
```

# Hovmöler Diagram


```{r image1, fig = TRUE, fig.width = 12, fig.height = 6}
# Default image plot
image(mod)
```

Customized plot which shows the "foehn frequency" for the
interesting time period from August to April with custom
colors and additional contour lines and custom
aggregation period (two-weeks, 3-hourly).


```{r image2, fig = TRUE, fig.width = 12, fig.height = 6}
# Customizing image plot
image(mod, deltad = 14L, deltat = 3*3600, contours = TRUE,
      contour.col = "white", lwd = 2, labcex = 1.5,
      col = colorspace::sequential_hcl(51, "Purple-Yellow", rev = TRUE),
      xlim = c(212, 119), zlim = c(0, 0.4))
```




