
```{r, echo = FALSE}
suppressPackageStartupMessages(library("foehnix"))
```

# Data Set Description

The "Californian" data set consists of hourly meteorological observations
from station "Viejas Casino and Ranch" and station "Lucky Five Ranch"
located in South California. 

Viejas is located at the foot of the westerly slope of the Sierra Nevada
mountain range and exhibits strong easterly winds during downslope wind
situations. The Lucky Five Ranch is located northeast of and provides
information about the upstream air mass for the classification algorithm.


```{r, echo = FALSE, fig.width = 10, fig.height = 5}
library("leaflet")
library("sp")

stations <- data.frame(lon  = c(-116.70437, -116.528),
                       lat  = c(32.84559, 32.9331),
                       alt  = c(715, 1445),
                       name = c("Viejas", "Lucky Five Ranch"),
                       stringsAsFactors = FALSE)

b <- list(x0 = min(stations$lon) - .2, x1 = max(stations$lon) + .2,
          y0 = max(stations$lat) + .2, y1 = max(stations$lat) + .2)

m <- leaflet(width = "100%") %>% addTiles() %>% fitBounds(b$x0, b$y0, b$x1, b$x2)
m <- setView(m, mean(stations$lon), mean(stations$lat), zoom = 10.5)
for (i in 1:nrow(stations)) {
    m <- addPopups(m, stations$lon[i], stations$lat[i],
                   sprintf("Station %s, %d m a.m.s.l.", stations$name[i],
                           stations$alt[i]))
}
m <- addProviderTiles(m, "OpenTopoMap")
m
```


# Loading the Data Set

[`demodata("california")`](references/demodata) returns a data set which
combines hourly meteorological observations of both sites (Viejas; Lucky Five).
In addition, the potential temperature difference between the two stations
is calculated by reducing the dry air temperature from "Lucky Five Ranch"
to the height of "Viejas" (dry adiabatic lapse rate of 1K per 100m;
stored on `diff_temp`).
For details see [`demodata`](references/demodata.html).

```{r load_california}
data <- demodata("california")
head(data, n = 3)
```

```{r, fig = TRUE, fig.width = 12, fig.height = 6}
par(mfrow = c(1,2))
# Observed wind speed/wind direction "Viejas"
windrose(data, ddvar = "wind_direction", ffvar = "wind_speed",
         main = "Windrose\nViejas Casino and Resort",
         breaks = seq(0, 22, by = 2))
# Observed wind speed/wind direction "Lucky Five"
windrose(data, ddvar = "crest_wind_direction", ffvar = "crest_wind_speed",
         main = "Windrose\nLucky Five Ranch", hue = 270,
         breaks = seq(0, 22, by = 2))
```

Given the plots above we define the `foehn` wind direction
at Viejas between 305 and 160 degrees (a `r 360 - (305 - 160)` degree
wind sector centered northeast). This wind sector can be chosen
rather wide, but should exclude non-foehn wind directions to exclude
upslope winds.

```{r, fig = TRUE, fig.width = 8, fig.height = 8}
windrose(data, ddvar = "wind_direction", ffvar = "wind_speed",
         main = "Windrose Viejas with Custom foehn Wind Sector",
         breaks = seq(0, 16, by = 2),
         windsector = list(wind_direction = c(305, 160)))
```

For the [`windrose`](references/windrose.html) plot the `windsector`
is solely used for visual justification, the same restriction will
be used in the following step when estimating the [`foehnix`](references/foehnix.html)
classification model.


# Estimate `foehnix` Model for Foehn Classification

The following model assumptions are used:

* Main variable to separate "foehn" and "no foehn" cases: `diff_temp` (potential temperature difference).
* Concomitant variable: `wind_speed` (wind speed at target station Viejas).
* Wind filter: the `wind_direction` at station Viejas has to lie within
    305 and 160 degrees (northeasterly wind direction; see above).
* Option `switch = TRUE` as high `diff_temp` indicate stable stratification (no foehn).

```{r estimate_model, results = "hide"}
mod    <- foehnix(diff_temp ~ wind_speed,
                  data   = data,
                  switch = TRUE,
                  filter = list(wind_direction = c(305, 160)))

```

### Model Summary

```{r, echo = FALSE}
mod
```

The full data set contains `N = `r nrow(mod$data)`` rows, 
`r nrow(data)` from the data set (`data`) and 
`r mod$inflated` due to inflation (create strictly regular time series object).
Thereof, `r length(mod$filter_obj$ugly)` are not considered
due to missing data, `r length(mod$filter_obj$bad)` as they do not fulfil the filter constraint
(`wind_direction` outside defined wind sector), wherefore the final
model is based on `r length(mod$filter_obj$good)` observations (or rows).


### Model coefficients

The following parameters are estimated for the two `r mod$control$family$name` clusters:

* No-foehn cluster: $\mu_1 = `r round(coef(mod)["mu1"], 2)`$,
                    $\sigma_1 = `r round(coef(mod)["sigma1"], 2)`$ (parameter scale)
* Foehn cluster: $\mu_2 = `r round(coef(mod)["mu2"], 2)`$,
                    $\sigma_2 = `r round(coef(mod)["sigma2"], 2)`$ (parameter scale)
* Concomitant model: `r ifelse(sign(coef(mod)["wind_speed"]), "positive", "negative")`
    `wind_speed` effect, `r sprintf("%+.1f", exp(coef(mod)["wind_speed"]) - 1)` percent per
    $m~s^{-1}$

```{r}
coef(mod)
```

Visual justification of the estimated clusters:

```{r plot_mod_hist, fig = TRUE, fig.width = 8, fig.height = 3.5}
plot(mod, which = "hist")
```

<div class="hide-Rinput">

```{r plot_mod_2d, fig = TRUE, fig.width = 8, fig.height = 6}
# Grid definition for 2-D plot
tmp <- list(diff_temp      = seq(-5, +12, by = .5),
            wind_speed     = seq(0, 12, by = .5),
            wind_direction = 45)
# Expand grid to create 'newdata' data.frame
grd  <- do.call(expand.grid, tmp)
mat  <- matrix(grd$diff_temp, nrow = length(tmp$diff_temp))
# Predict probabilities
pred <- matrix(predict(mod, newdata = grd)$prob, nrow = length(tmp$diff_temp))
# Create plot
cols <- colorspace::diverging_hcl(51, p1 = 5)
par(mar = rep(.5, 4), oma = c(3.1, 3.1, 2.1, 3.1))
layout(matrix(1:2, nrow = 1L), widths = c(10, 1))
image(x = tmp$diff_temp, y = tmp$wind_speed, z = pred, col = cols, zlim = c(0,1))
mtext(side = 1, line = 2.5, "diff_temp")
mtext(side = 2, line = 2.5, "wind_speed")
image(matrix(seq(0, 1, length = 51), nrow = 1),
      col = cols, yaxt = "n", xaxt = "n")
mtext(side = 4, line = 2.5, "probability")
mtext(side = 3, outer = TRUE, font = 2, cex = 1,
      "Predicted Probability | Wind Direction = 45")
axis(side = 4, at = seq(0, 1, by = .1))
```

</div>



# Graphical Model Assessment

A [`foehnix`](references/foehnix.html) object comes with generic plots for graphical model
assessment. The following figure shows the
'log-likelihood contribution' of

* the main **component** (left hand side of formula),
* the **concomitant** model (right hand side of formula),
* and the **full** log-likelihood sum which is maximised by
  the optimization algorithm.

The abscissa shows the logarithm of the iterations during 
optimization. 

```{r plot_mod1, fig = TRUE, fig.width = 12, fig.height = 6}
# Log-likelihood contribution
plot(mod, which = "loglikcontribution")
```

Alternatively the coefficient paths can be plotted:

```{r plot_mod2, fig = TRUE, fig.width = 12, fig.height = 6}
# Coefficient path
plot(mod, which = 3L)
```

The left plot shows the parameters of the two components
($\mu_1$, $\log(\sigma_1)$, $\mu_2$, $\log(\sigma_2)$), the
right one the standardized coefficients of the concomitant model.


# Time Series Plot

The Californian demo data set has non-standard variable names (by purpose).
Thus, when calling `tsplot` (time series plot) we do have to manually specify
these names.

```{r, fig = TRUE, fig.width = 12, fig.height = 15}
# Some smaller quality issues in the data (should not be a big deal)
start <- as.POSIXct("2012-03-01")
end   <- as.POSIXct("2012-03-12")

# As we dont have the standard names: re-specify the
# names for this plot (new feature).
tsplot(mod, style = "advanced", diff_t = "diff_temp",
       t   = "air_temp",         crest_t   = "crest_air_temp",
       dd  = "wind_direction",   crest_dd  = "crest_wind_direction",
       ff  = "wind_speed",       crest_ff  = "crest_wind_speed",
       ffx = "wind_gust",        crest_ffx = "crest_wind_gust",
       rh = "relative_humidity",
       windsector = list(wind_direction = c(305, 160)),
       start = start, end = end)
```


# Wind Rose Plot

```{r windrose}
windrose(mod, dd = "wind_direction", ff = "wind_speed",
         type = "hist", which = c("foehn", "nofoehn"),
         windsector = list(c(305, 160)))
```

# Hovmöler Diagram


```{r image_1, fig = TRUE, fig.width = 12, fig.height = 6}
# Default plot
image(mod)
```

Customized plot which shows the "foehn frequency" for the
interesting time period from August to April with custom
colors and additional contour lines and custom
aggregation period (two-weeks, 3-hourly).


```{r image_2, fig = TRUE, fig.width = 12, fig.height = 6}
image(mod, deltad = 14L, deltat = 3*3600, contours = TRUE,
      contour.col = "white", lwd = 2, labcex = 1.5,
      col = colorspace::sequential_hcl(51, "Purple-Yellow", rev = TRUE),
      xlim = c(212, 119), zlim = c(0, 0.4))
```




