
```{r, echo = FALSE}
suppressPackageStartupMessages(library("foehnix"))
```



```{r, echo = FALSE, fig.width = 10, fig.height = 5}
library("leaflet")
library("sp")

stations <- data.frame(lon  = c(-116.70437, -116.528),
                       lat  = c(32.84559, 32.9331),
                       alt  = c(715, 1445),
                       name = c("Viejas", "Lucky Five Ranch"),
                       stringsAsFactors = FALSE)

b <- list(x0 = min(stations$lon) - .2, x1 = max(stations$lon) + .2,
          y0 = max(stations$lat) + .2, y1 = max(stations$lat) + .2)

m <- leaflet() %>% addTiles() %>% fitBounds(b$x0, b$y0, b$x1, b$x2)
m <- setView(m, mean(stations$lon), mean(stations$lat), zoom = 10.5)
for (i in 1:nrow(stations)) {
    m <- addPopups(m, stations$lon[i], stations$lat[i],
                   sprintf("Station %s, %d m a.m.s.l.", stations$name[i],
                           stations$alt[i]))
}
m <- addProviderTiles(m, "OpenTopoMap")
m
```

# Loading the Data Set

[`demodata("california")`](references/demodata) returns a data set which
combines hourly meteorological observations for the two sites
"Viejas Casino and Resort" and "Lucky Five Ranch" which serves as the
crest station.

In addition, the potential temperature difference between the two stations
is calculated by reducing the dry air temperature from "Lucky Five Ranch"
to the height of "Viejas" (dry adiabatic lapse rate of 1K per 100m;
stored on `diff_t`).

```{r load_california}
data <- demodata("california")
head(data, n = 3)
```

# Estimate `foehnix` Model for Foehn Classification

We will use the following model assumptions:

* Main variable to separate "foehn" and "no foehn" cases: `diff_t` (potential temperature difference).
* Concomitant variable: `wind_speed` (wind speed at target station Viejas).
* Wind filter: the `wind_direction` at station Viejas has to lie within
    305 and 160 degrees (easterly wind direction; downslope).
* Option `switch = TRUE` as high `diff_temp` indicate stable stratification (no foehn).

```{r estimate_model, results = "hide"}
mod    <- foehnix(diff_temp ~ wind_speed,
                  data   = data, switch = TRUE,
                  filter = list(wind_direction = c(305, 160)))

```

Which gives us the following model:

```{r}
mod
```


# Time Series Plot

The Californian demo data set has non-standard variable names (by purpose).
Thus, when calling `tsplot` (time series plot) we do have to manually specify
these names.

```{r, fig = TRUE, fig.width = 12, fig.height = 15}
# Some smaller quality issues in the data (should not be a big deal)
start <- as.POSIXct("2012-03-01")
end   <- as.POSIXct("2012-03-12")

# As we dont have the standard names: re-specify the
# names for this plot (new feature).
tsplot(mod, style = "advanced", diff_t = "diff_temp",
       t   = "air_temp",         crest_t   = "crest_air_temp",
       dd  = "wind_direction",   crest_dd  = "crest_wind_direction",
       ff  = "wind_speed",       crest_ff  = "crest_wind_speed",
       ffx = "wind_gust",        crest_ffx = "crest_wind_gust",
       rh = "relative_humidity",
       windsector = list(wind_direction = c(305, 160)),
       start = start, end = end)
```

### Model coefficients

```{r}
coef(mod)
```


# Graphical Model Assessment

```{r plot_mod, fig = TRUE, fig.width = 12, fig.height = 6}
# Log-likelihood contribution
plot(mod, which = 2L)
# Coefficient path
plot(mod, which = 3L)
# Histogram
plot(mod, which = 4L)
```

# Wind Rose Plot

```{r windrose}
windrose(mod, dd = "wind_direction", ff = "wind_speed",
         type = "hist", which = c("foehn", "nofoehn"),
         windsector = list(c(305, 160)))
```

# Hovmöler Diagram


```{r image_1, fig = TRUE, fig.width = 12, fig.height = 6}
# Default plot
image(mod)
```

Customized plot which shows the "foehn frequency" for the
interesting time period from August to April with custom
colors and additional contour lines and custom
aggregation period (two-weeks, 3-hourly).


```{r image_2, fig = TRUE, fig.width = 12, fig.height = 6}
image(mod, deltad = 14L, deltat = 3*3600, contours = TRUE,
      contour.col = "white", lwd = 2, labcex = 1.5,
      col = colorspace::sequential_hcl(51, "Purple-Yellow", rev = TRUE),
      xlim = c(212, 119), zlim = c(0, 0.4))
```




