
```{r, echo = FALSE}
suppressPackageStartupMessages(library("foehnix"))
```

## Windrose

The `foehnix` package comes with methods to create windrose plot for
foehn classification models (see [getting started](foehnix.html),
[`foehnix reference`](../references/foehnix.html)) and observation
data. Two types of windrose plots are available:

* **density**: empirical density
* **histogram**: empirical circular histogram

### Windrose Plot for Observation Data

The [`windrose`](../references/windrose.html) function
allows to plot empirical circular densities and histograms (the classical
'windrose') of observed values. The data set can either be a

* multivariate `zoo` time series object,
* a `data.frame`,
* or simply two numeric vectors.

All needed is wind speed ($\in [0, \infty]$) and wind direction.
The wind direction needs to be provided in meteorological degrees
($\in [0, 360]$) where $0$ and $360$ corresponds to wind coming from North,
$90$ for wind from East, $180$ for wind from South, and $270$ from
West.

If used with a multivariate `zoo` object or a `data.frame` the
[`windrose`](../references/windrose.html) function expects to find two
variables called `ff` (wind speed) and `dd` (wind direction).
However, custom names can be specified if needed (see customization section).
The plot below shows examples using `zoo`, `data.frame`, or `numeric` vectors.
While the upper two show `type = "density"`, the lower two show the circular
histograms (`type = "histogram"`).



```{r, echo = TRUE, figure = TRUE, fig.width = 8, fig.height = 8}
# Loading the demo data set for station Ellboegen
data <- demodata("ellboegen")
print(head(data))
class(data)

# Default plot using a zoo object
par(mfrow = c(2, 2), mar = c(1, 1, 3, 1))
windrose(data)
# Default plot using a data.frame
windrose(data.frame(data))
# Defualt plot using univariate zoo objects/numeric vectors
windrose(data$dd, data$ff, type = "histogram")
windrose(as.numeric(data$dd), as.numeric(data$ff), type = "histogram")
```

### Windrose Plot for foehnix Models

The [`windrose`](../references/windrose.html) function can also directly
be applied to [`foehnix`](../references/foehnix.html) objects.
By default, six windroses will be plotted:

* **top row:** density plot
* **bottom row:** circular histograms
* **left to right:** unconditional, for $\pi < 0.5$ and $\pi \ge 0.5$

_Unconditional_ is the same as if one would call the [`windrose`](../references/windrose.html)
function on the data set (`data`), the latter two show the windrose
conditional on the foehn probability estimated by the
[`foehnix`](../references/foehnix.html) classification model.

```{r, echo = TRUE, figure = TRUE, fig.width = 8, fig.height = 6.5}
# Loading the demo data set for Tyrol (Ellboegen and Innsbruck)
data <- demodata("tyrol")

# Estimate a foehnix classification model
filter <- list(dd = c(43, 223), crest_dd = c(90, 270))
mod <- foehnix(diff_t ~ ff + rh, data = data, filter = filter,
               switch = TRUE, verbose = FALSE)

# Plotting windroses
windrose(mod)
```

Again, the [`windrose`](../references/windrose.html) function expects
that the two variables 'wind speed' and 'wind direction' are called
`ff` and `dd` but can be adjusted if custom names are used.
An example:

```{r windrose_foehnix, echo = TRUE, figure = TRUE, fig.width = 8, fig.height = 6.5}
# Loading the demo data set for station Ellboegen and Sattelberg (combined)
data <- demodata("tyrol") # default
names(data) <- gsub("dd$", "winddir", names(data))
names(data) <- gsub("ff$", "windspd", names(data))
names(data)

# Estimate a foehnix classification model using the new
# custom names (see 'foehnix' function documentation for details)
filter <- list(winddir = c(43, 223), crest_winddir = c(90, 270))
mod2 <- foehnix(diff_t ~ windspd + rh, data = data, filter = filter,
                switch = TRUE, verbose = FALSE)

# Plotting windroses using custom names
windrose(mod2, ddvar = "winddir", ffvar = "windspd")
```

The additional input arguments `type` and `which` allow to 
specify what should be plotted:

```{r windrose_foehnix_subset, echo = TRUE, figure = TRUE, fig.width = 8, fig.height = 5}
# density plots only
windrose(mod, type = "density", ncol = 3)
# circular histogram plots only
windrose(mod, type = "histogram", ncol = 3)
# Only histograms for "foehn" and "no foehN"
windrose(mod, type = "histogram", which = c("nofoehn", "foehn"))
```


# Windrose customization









